---
title: "Danish kings assignment"
author: "Sofie, Kristiane og Emilie"
date: "Oprettet 7. marts 2025, opdateret 24 maj 2025"
output:
  html_document:
    toc: true         
    toc_depth: 2  
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library,echo=FALSE}
library(tidyverse)

read_csv2("data/danish_monarchs_clean.csv",na = "NA")
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: the seperator is semicolon (;)  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/danish_monarchs_clean.csv")
glimpse(kings1)
kings2 <- read_csv("data/danish_monarchs_clean.csv")
glimpse(kings2)
kings3 <- read.csv2("data/danish_monarchs_clean.csv")
glimpse(kings3)
kings4 <- read_csv2( "data/danish_monarchs_clean.csv")
glimpse(kings4)

```
Answer: 
1. Which of these functions is a `tidyverse` function? 
read_csv2 is a 'tidyverse' function.

2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
The result is: [1] "spec_tbl_df" "tbl_df"      "tbl"         "data.frame" 

3. How many columns does the object have when created with these different functions?
We use the function: ncol(). There is 14 columns.

4. Show the dataset so that we can see how R interprets each column.
We use the function: glimpse()

```{r }
class(kings1)
class(kings2)
class(kings3)
class(kings4)
```
```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- kings4

class(kings)


ncol(kings)


glimpse(kings)


```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

We already have a column that shows the duration of reign called 'Periode'. But we will show below how we use the mutate function to create a new column called 'duration' in the kings dataset.

```{r}
# YOUR CODE

kings <- kings %>%
  mutate(duration = Slut_regeringsaar - Start_regeringsaar)

glimpse(kings)

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* If your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`

As seen below we have tried to use both the mean() base-R function and the`summarize()` tidyverse function.
Both functions give us the same result: The average duration of reign of all rulers is 19.9 years.


```{r}
# YOUR CODE
kings_duration <- c(kings$duration)
kings_duration

mean(kings_duration, na.rm = TRUE)

kings_duration_tibble <- kings %>%
  select(duration)
kings_duration_tibble

kings_duration_tibble %>%
  summarise(mean_duration = mean(duration, na.rm = TRUE))

```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

By using the code below, we have come to the conclusion that 25 kings enjoyed a longer-than-averiage duration of reign.
Under Step 2 we have created a tibble with the names of the 25 kings.

```{r}
# YOUR CODE
# Step 1: Calculate the average duration
mean_duration <- mean(kings$duration, na.rm = TRUE)

# Step 2: Filter kings with a longer-than-average duration
longer_than_average_kings <- kings %>%
  filter(duration > mean_duration) %>% 
  select(Navn)

print(longer_than_average_kings)

# Step 3: Count the resulting rows
longer_than_average_kings_count <- longer_than_average_kings %>%
  count()

# Step 4: Display the count
print(longer_than_average_kings_count)
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

The three longest-ruling monarchs is:
Christian 4.: 60 years or 21901 days.
Margrethe 2.: 52 years or 18981
Erik 7. Pommern: 43 years or 15695 days. 

```{r}
# YOUR CODE
# Step 1: Sort kings by duration in descending order and select top 3
top_3_kings <- kings %>%
  arrange(desc(duration)) %>%
  slice(1:3)
top_3_kings

top_3_kings_with_days <- top_3_kings %>% 
  mutate(Days = duration * 365 + ifelse(Periode %% 4 == 0 & (Periode %% 100 != 0 | Periode %% 400 == 0), 1, 0))
  
print(top_3_kings_with_days)

```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ?
See the answer in danish_kings_assignment.pdf in the GitHub repository.

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
# Step 1: Create the 'midyear' column
kings <- kings %>%
  mutate(midyear = Start_regeringsaar + (Slut_regeringsaar - Start_regeringsaar) / 2)

glimpse(kings)

# Step 2: Plot the duration of reign with ggplot
duration_of_reign_danish_monarchs <- ggplot(kings,aes(x=midyear,y=duration)) + 
  geom_point() + 
  geom_smooth() + 
  labs(
    title = "Duration of Reign of Danish Monarchs Through Time",
    x = "Year (Midpoint of Reign)",
    y = "Duration of Reign (years)"
  ) + 
  theme_bw() +
theme(
    plot.title = element_text(size = 20, face = "bold"),       
    axis.title.x = element_text(size = 18),                    
    axis.title.y = element_text(size = 18),                    
    axis.text.x = element_text(size = 18),                     
    axis.text.y = element_text(size = 18)                     
  )


# Step 3: See the result
duration_of_reign_danish_monarchs
  

ggsave("figures/duration_of_reign_danish_monarchs.png", duration_of_reign_danish_monarchs , width = 15, height = 10)
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


